Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-24 draft/draft_v3.tex
more draft

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
\author[1,*]{Juan Diaz-Colunga}
\author[1,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[1]{Chang-Yu Chang}
\author[1]{Helen S. Cai}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,\Letter]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

\affil[$\textrm{\Letter}$]{\normalfont alvaro.sanchez@yale.edu}
\affil[*]{\normalfont These authors contributed equally}
\date{}

@@

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
We carried out our simulations enabling a large variation across species in terms of
their metabolic architectures;
this choice is discussed in \hyperref[box1]{Box~1}.
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive and resident communities respectively,
that were allowed to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to simulate our coalescence
and dominant-dominant competition experiments
(\methodsref{sim}).
We found that the MicroCRM was able to capture a correlation between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[b]{fig2}, right panel),
further supporting the idea that top-down ecological co-selection can consistently
emerge from metabolic interactions across species.

@@

\begin{mdframed}
\internallinenumbers
\subsubsection*{Box 1: Community cohesiveness in a Microbial Consumer-Resource Model}
\label{box1}

The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
is a modeling framework based on the classic MacArthur's consumer resource model
\cite{MacArthur1970}.
It encodes the dynamics of a system with $S$ species and $M$ resources
in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
with an additional set of parameters controlling
the species maintenance costs ($m_i$ for species $i$),
the resource energy densities ($w_\alpha$ for resource $\alpha$),
the energy to growth rate conversion factor ($g_i$ for species $i$)
or the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
($l_\alpha$ for resource $\alpha$).
The element $c_{i\alpha}$ of the consumer preference matrix
represents the uptake rate of resource $\alpha$ by
species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
complex in modeling scenarios that are not considered here,
see \cite{Goldford2018,Marsland2019,Marsland2020}).
The element $D_{\beta\alpha}$ of the metabolic matrix
represents the amount of energy secreted in the form of resource $\beta$
as a result of the metabolism of resource $\alpha$.

This formulation assumes that all species metabolize resources through similar pathways,
with any two species secreting the same byproducts when consuming a same resource.
However, experimental evidence suggests that individual species can secrete different sets of metabolites
to the environment when growing on a same primary resource
\cite{Harcombe2014,Pinu2018,Estrela2020}. This observation motivated us to introduce a new
feature in the model:
in short, we now define $\mathbf{D}$ as a three-dimensional matrix
where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
that is secreted \textit{by species $i$} when it metabolizes resource $\alpha$.
Note that now $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
(see illustration below).

@@

@@

Our data indicates that the primary resource supplied to the communities can modulate
the effect of the cohorts in the dominants pairwise competition
(\figref[a]{fig2})
and the strength of top-down co-selection
(\figref[b]{fig2}, left and middle panels).
Our model suggests that this might be a result of the metabolic interactions
between community members, including the rarer taxa
(\hyperref[box1]{Box~1}).
To investigate the potential role of the cohorts in coalescence,
 i.e. whether the dominants may be co-selected for or against by them
(\figref[e]{fig1}, right panels),
we ran a new set of simulations
this time invading resident communities with the dominants alone
(\methodsref{sim}).
We compared the invasion success of the dominants in isolation with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominants was quantified by their relative abundance
in the final stabilized communities.
Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves
(\figref[b]{fig3}, green shaded region).
On the other hand, a high degree of antagonistic bottom-up co-selection would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).

\figref[b]{fig3} shows that, in the simulations,
many dominants could not invade on their own
(or could only do so at very low final relative abundances, below 0.1)
but were able to reach high frequencies when they were accompanied by their cohorts.
This indicates that positive bottom-up co-selection is frequent and potentially very strong,
while negative bottom-up co-selection is far more uncommon.
We then asked whether the ability of the pairwise competition of dominants
to predict coalescence outcomes was dependent on the strength of bottom-up co-selection.
We divided our simulations into two subsets:
the first one was comprised of the instances where positive bottom-up co-selection was strong
(i.e. dots in the green shaded region of \figref[b]{fig3}),
the second set included all other cases
(dots near the diagonal of \figref[b]{fig3}).
We went back to our original simulations and plotted the frequency of the invasive dominant
in pairwise competition with the resident dominant versus the relative similarity between
the invasive and coalesced communities,
i.e. the same plot as in \figref[b]{fig2}, for each subset.
We found that when bottom-up positive co-selection is strong,
the pairwise competition of dominants is not predictive of coalescence outcomes
(\figref[c]{fig3}, left panel) and vice-versa (\figref[c]{fig3}, right panel).

We then asked whether this trend was also observed \textit{in vitro}.
We went back to our synthetic communities
and carried out a new round of experiments where we invaded the resident communities
with the invasive dominants alone
(\methodsref{competitions}).
After stabilization, we quantified species abundance through 16S Illumina sequencing
(\methodsref{sequencing}).
Again, we observed that bottom-up co-selection is far more common in its
positive than in its negative form
(\figref[d]{fig3}).
Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
environments than in the citrate ones,
consistent with our hypothesis that metabolic interactions among species
are key in determining the strength and direction of ecological co-selection.
We then repeated our analysis in \figref[c]{fig3},
this time splitting our data according to the observed strength of bottom-up
co-selection instead of the primary carbon source
as we had done in \figref[b]{fig2}.
Our findings were in line with the model prediction:
pairwise competition between dominants is only predictive of coalescence outcomes
if bottom-up co-selection is weak.

@@

@@

Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic networks that shape the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
This network is in turn regulated by the supplied primary carbon source.
This idea is supported by the observation that a Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model also predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

@@

Beta-diversity indexes between the invasive and coalesced communities
or the resident and coalesced communities were computed using various
similarity metrics. For two arbitrary communities with
ESV abundances represented by the vectors
$\mathbf{x} = \left( x_1, x_2, \cdots, x_N \right)$
and
$\mathbf{y} = \left( y_1, y_2, \cdots, y_N \right)$
(where $x_i$ and $y_i$ represent the relative abundance of the $i$th ESV in
each community respectively and $N$ is the total number of ESVs), the Bray-Curtis
similarity $BC \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
\cite{Bray1957}

@@

Experimental data and code for the analysis, as well as code for the simulations
and the updated Community Simulator package with instructions for enabling the
new features are deposited in \url{github.com/jdiazc9/coalescence}.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=10cm,keepaspectratio]{figs/fig1.pdf}
\caption{\textbf{Overview of the experimental protocol.}
\textbf{a.}~Environmental samples collected from eight different locations were used
to inoculate our communities.
\textbf{b.}~Communities were stabilized in serial batch culture bioreactors
\cite{Goldford2018} in minimal synthetic media with glutamine or citrate as the
only supplied carbon source.
\textbf{c.}~Communities were plated in minimal media agar plates and the most abundant
species (the ``dominants") from each community were isolated. We refer to the set of
sub-dominant species as the ``cohorts".
\textbf{d.}~Rank-frequency distributions of all eight communities stabilized in either
glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
Three biological replicates per community are shown.
Community compositions are skewed and long-tailed.
\textbf{e.}~Our hypothesis is that ecological co-selection can take place from the top-down,
i.e. the dominant co-selecting the cohort, or from the bottom-up, i.e. the cohort co-selecting
the dominant. Both forms of co-selection can be positive (recruitment) or negative
(antagonism).
\textbf{f.}~Illustration of the protocol of our coalescence experiments. All pairs of
communities were inoculated into fresh minimal media supplemented with the same carbon
source where communities had been previously stabilized. The coalesced (C) and original
resident (R) and invasive (I) communities were then serially diluted and allowed to grow
for seven additional transfers.}
\label{fig1}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
\caption{\textbf{Top-down co-selection in microbial community coalescence.}
\textbf{a.} Pairwise competition of dominants with or without their cohorts.
In the horizontal axis, we plot the frequency of the invasive dominant species
in head-to-head pairwise competition with the resident dominant.
In the vertical axis, we plot the same relative frequency when the two species compete
in the presence of their cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine (red)
and $R^2=0.83$, $p<10^{-8}$ for citrate (blue).
\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species.
Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
A high correlation is consistent with a scenario of
strong top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
Right panel (black): simulations with a Microbial Consumer-Resource Model are able
to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
\label{fig2}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
or they could have no effect in the invasion success of the dominant taxa (gray).
\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
We plot the frequency reached by the invasive dominants when invading the resident communities
in isolation versus the same frequency when invading together with their cohorts,
i.e. in community coalescence.
Points in the green/red area represent instances where the invasive dominant is able to invade
with higher/lower success when accompanied by its cohort, evidencing positive/negative
bottom-up co-selection.
Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
is only weakly affected by the presence or absence of its cohort.
\textbf{c.} We divided the data from our simulations into two sets according to whether
positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
is poorly predictive of coalescence outcomes
($R^2=0.00$, $p>0.05$).
Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
strongly dictated by the result of the dominant-dominant competition
($R^2=0.34$, $p<10^{-6}$).
\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
frequent and strong than negative bottom-up co-selection.
\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
instead of by the carbon source provided to the communities.
Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
yield a poor correspondence between dominant pairwise competition and coalescence outcomes
(left panel, $R^2=0.07$, $p>0.05$)
and vice-versa
(right panel, $R^2=0.37$, $p<10^{-4}$).}
\label{fig3}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=13cm,keepaspectratio]{figs/fig4_v2.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...}
\label{fig4}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=13cm,keepaspectratio]{figs/fig5.pdf}
\caption{\textbf{Community hierarchies determine coalescence outcomes.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...
\textbf{b.} Blablabla... ($p=0.083$).}
\label{fig4}
\end{figure}

>>>>

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
\author[1 *]{Juan Diaz-Colunga}
\author[1 *]{Nanxi Lu}
\author[1,2 *]{Alicia Sanchez-Gorostiaga}
\author[1]{Chang-Yu Chang}
\author[1]{Helen S. Cai}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1 \Letter]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

\affil[$\textrm{\Letter}$]{\normalfont~alvaro.sanchez@yale.edu}
\affil[*]{\normalfont~These authors contributed equally}
\date{}

@@

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
for Python \cite{Marsland2020}
(\hyperref[box1]{Box~1}).
We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
Based on previous experimental work
\cite{Harcombe2014,Pinu2018,Estrela2020},
we allowed different species in our model
to secrete different sets of byproducts when metabolizing a same resource
(\hyperref[box1]{Box~1}).
\iffalse
this choice is discussed in \hyperref[box1]{Box~1}.
\fi
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive
and 100 resident communities respectively
by randomly choosing 50 species and
allowing them to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to simulate our coalescence
and dominant-dominant competition experiments
(\methodsref{sim}).
We found that the MicroCRM was able to capture a correlation between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[b]{fig2}, right panel),
further supporting the idea that top-down ecological co-selection can consistently
emerge from metabolic interactions across species.

@@

\begin{mdframed}
\internallinenumbers
\subsubsection*{Box 1: A Microbial Consumer-Resource Model for community coalescence}
\label{box1}

The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
is a modeling framework based on the classic MacArthur's consumer resource model
\cite{MacArthur1970}.
It encodes the dynamics of a system with $S$ species and $M$ resources
in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
with an additional set of parameters controlling
the species maintenance costs ($m_i$ for species $i$),
the resource energy densities ($w_\alpha$ for resource $\alpha$),
the energy to growth rate conversion factor ($g_i$ for species $i$)
and the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
($l_\alpha$ for resource $\alpha$).
The element $c_{i\alpha}$ of the consumer preference matrix
represents the uptake rate of resource $\alpha$ by
species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
complex in modeling scenarios that are not considered here,
see \cite{Goldford2018,Marsland2019,Marsland2020}).
Experimental evidence suggests that individual species can secrete different sets of metabolites
to the environment when growing on a same primary resource
\cite{Harcombe2014,Pinu2018,Estrela2020}.
Thus, we define $\mathbf{D}$ as a three-dimensional matrix
where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
that is secreted by species $i$ when it metabolizes resource $\alpha$.
Note that $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
(see illustration below).
\iffalse
For our simulations,
the elements of $\mathbf{c}$ are sampled from a gamma distibution,
while the elements of $\mathbf{D}$ are sampled from a Dirichlet distribution
to ensure that the normalization condition $\sum_\beta D_{i\beta\alpha} = 1$
is satisfied for every $i$ and $\alpha$.
\fi
\iffalse
The sampling is performed so that species belonging to a same family
tend to secrete more similar byproducts when consuming a same resource
(see \methodsref{sim} and \cite{Marsland2019,Marsland2020}),
as supported by experimental observations
\cite{Harcombe2014,Pinu2018,Estrela2020}.
\fi

@@

The following equations describe the kinetics of the abundances
of the $i$-th species (denoted as $N_i$)
and the $\alpha$-th resource (denoted as $R_\alpha$):

\begin{equation}
\frac{dN_i}{dt} = 
g_i N_i
\left[
\sum_\alpha \left( 1-l_\alpha \right)
w_\alpha c_{i\alpha} R_\alpha
- m_i
\right]
\label{eq:crm-n}
\end{equation}

\begin{equation}
\frac{dR_\alpha}{dt} = 
- \sum_j N_j c_{j\alpha} R_\alpha
+ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
\left[
l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
\right]
\label{eq:crm-r}
\end{equation}
%
These equations can take slightly different forms in certain cases,
e.g. if the primary resource is supplied continuously instead of at the beggining
of each growth cycle \cite{Marsland2019,Marsland2020},
but they are a good approximation for the community dynamics between consecutive
serial dilutions in our setup.
Here, we assembled \textit{in silico} communities by randomly
sampling a set of species from a pool,
then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
diluting the final abundances,
replenishing the primary resource (by default denoted as $R_0$),
and repeating the process until generational equilibrium was achieved
(\methodsref{sim}).
Coalescence simulations were carried out following the same logic,
this time seeding the coalesced communities by mixing the invasive and resident
communities instead of sampling from a species pool.

@@

Our data indicates that the primary resource supplied to the communities can modulate
the effect that the cohorts have in the dominants pairwise competition
(\figref[a]{fig2})
and the strength of top-down co-selection
(\figref[b]{fig2}, left and middle panels).
The fact that
our model captures these trends
suggests that this might be a result of the metabolic interactions
between community members, including the rarer taxa.
To investigate the potential role of the cohorts in coalescence,
 i.e. whether the dominants may be co-selected for or against by them
(\figref[e]{fig1}, right panels),
we ran a new set of simulations
this time invading resident communities with the dominants alone
(\methodsref{sim}).
We compared the invasion success of the dominants in isolation with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominants was quantified by their relative abundance
in the final stabilized communities.
Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves
(\figref[b]{fig3}, green shaded region).
On the other hand, a high degree of bottom-up antagonism would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).

\figref[b]{fig3} shows that, in the simulations,
many dominants could not invade on their own
(or could only do so at very low final relative abundances, below 0.1)
but were able to reach high frequencies when they were accompanied by their cohorts.
This indicates that positive bottom-up co-selection is frequent and potentially very strong,
while negative bottom-up co-selection is far more uncommon.
We then asked whether the ability of the pairwise competition of dominants
to predict coalescence outcomes was dependent on the strength of bottom-up co-selection.
We divided our simulations into two subsets:
the first one was comprised of the instances where positive bottom-up co-selection was strong
(i.e. dots in the green shaded region of \figref[b]{fig3}),
the second set included all other cases
(dots near the diagonal of \figref[b]{fig3}).
We reexamined our original simulations and plotted the frequency of the invasive dominant
in pairwise competition with the resident dominant versus the relative similarity between
the invasive and coalesced communities,
i.e. the same plot as in \figref[b]{fig2}, for each subset.
We found that when bottom-up positive co-selection is strong,
the pairwise competition of dominants is not predictive of coalescence outcomes
(\figref[c]{fig3}, left panel) and vice-versa (\figref[c]{fig3}, right panel).

We then asked whether this trend was also observed \textit{in vitro}.
We went back to our laboratory cultures
and carried out a new round of experiments where we invaded the resident communities
with the invasive dominants alone
(\methodsref{competitions}).
After stabilization, we quantified species abundance through 16S Illumina sequencing
(\methodsref{sequencing}).
Again, we observed that bottom-up co-selection is far more common in its
positive than in its negative form
(\figref[d]{fig3}).
Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
environments than in the citrate ones,
consistent with our hypothesis that metabolic interactions among species
are key in determining the strength and direction of ecological co-selection.
We then repeated our analysis in \figref[c]{fig3},
this time splitting our data according to the observed strength of bottom-up
co-selection instead of the primary carbon source
as we had done in \figref[b]{fig2}.
Our findings were in line with the model prediction:
pairwise competition between dominants is only predictive of coalescence outcomes
if bottom-up co-selection is weak
(\figref[e]{fig3}, $R^2=0.07$, $p>0.05$ when bottom-up co-selection is strong;
$R^2=0.37$, $p<10^{-4}$ when bottom-up co-selection is weak).

@@

\subsubsection*{Understanding the mechanisms of ecological co-selection: a minimal model of community coalescence}

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection,
we developed a minimal model of community coalescence.
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
As a result of $\mathrm{R}_1$ metabolism,
the dominants secrete a single byproduct off which
the sub-dominant species ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource that can again
be metabolized by the respective dominants.
This is the simplest structure where both top-dwn and bottom-up
cohesiveness can emerge, with their relative strengths being determined
by how effectively the species can utilize each other's metabolic byproducts.
Several parameters can modulate the direction of community cohesiveness,
and even in this simple model there can be complex interactions between them.
For instance, a sub-dominant that is very efficient at metabolizing the dominant
secretions will therefore be a strong producer of secondary secretions,
in turn increasing the metabolic flow towards the dominant itself.

\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selction}

How do the ideas above scale to more complex and diverse communities?

@@

Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic networks that shape the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
This network is in turn regulated by the supplied primary carbon source
in our minimal laboratory conditions.
This idea is supported by the observation that a Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model also predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

@@

Beta-diversity indexes between the invasive and coalesced communities
or the resident and coalesced communities were computed using various
similarity metrics. For two arbitrary communities with
ESV abundances represented by the vectors
$\mathbf{x} = \left( x_1, x_2, \cdots, x_S \right)$
and
$\mathbf{y} = \left( y_1, y_2, \cdots, y_S \right)$
(where $x_i$ and $y_i$ represent the relative abundance of the $i$th ESV in
each community respectively and $S$ is the total number of ESVs), the Bray-Curtis
similarity $BC \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
\cite{Bray1957}

@@

Experimental data and code for the analysis, as well as code for the simulations
and the updated Community Simulator package with instructions for enabling the
new features are in \url{github.com/jdiazc9/coalescence}.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig1.pdf}
\caption{\textbf{Overview of the experimental protocol.}
\textbf{a.}~Environmental samples collected from eight different locations were used
to inoculate our communities.
\textbf{b.}~Communities were stabilized in serial batch culture bioreactors
\cite{Goldford2018} in minimal synthetic media with glutamine or citrate as the
only supplied carbon source.
\textbf{c.}~Communities were plated in minimal media agar plates and the most abundant
species (the ``dominants") from each community were isolated. We refer to the set of
sub-dominant species as the ``cohorts".
\textbf{d.}~Rank-frequency distributions of all eight communities stabilized in either
glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
Three biological replicates per community are shown.
Community compositions are skewed and long-tailed.
\textbf{e.}~Our hypothesis is that ecological co-selection can take place from the top-down,
i.e. the dominant co-selecting the cohort, or from the bottom-up, i.e. the cohort co-selecting
the dominant. Both forms of co-selection can be positive (recruitment) or negative
(antagonism).
\textbf{f.}~Illustration of the protocol of our coalescence experiments. All pairs of
communities were inoculated into fresh minimal media supplemented with the same carbon
source where communities had been previously stabilized. The coalesced (C) and original
resident (R) and invasive (I) communities were then serially diluted and allowed to grow
for seven additional transfers.}
\label{fig1}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig2_v3.pdf}
\caption{\textbf{Top-down co-selection in microbial community coalescence.}
\textbf{a.} Pairwise competition of dominants with or without their cohorts.
In the horizontal axis, we plot the frequency of the invasive dominant species
in head-to-head pairwise competition with the resident dominant.
In the vertical axis, we plot the same relative frequency when the two species compete
in the presence of their cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine (red)
and $R^2=0.83$, $p<10^{-8}$ for citrate (blue).
\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species.
Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
A high correlation is consistent with a scenario of
strong top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
Right panel (black): simulations with a Microbial Consumer-Resource Model are able
to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
\label{fig2}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
or they could have no effect in the invasion success of the dominant taxa (gray).
\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
We plot the frequency reached by the invasive dominants when invading the resident communities
in isolation versus the same frequency when invading together with their cohorts,
i.e. in community coalescence.
Points in the green/red area represent instances where the invasive dominant is able to invade
with higher/lower success when accompanied by its cohort, evidencing positive/negative
bottom-up co-selection.
Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
is only weakly affected by the presence or absence of its cohort.
\textbf{c.} We divided the data from our simulations into two sets according to whether
positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
is poorly predictive of coalescence outcomes
($R^2=0.00$, $p>0.05$).
Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
strongly dictated by the result of the dominant-dominant competition
($R^2=0.34$, $p<10^{-6}$).
\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
frequent and strong than negative bottom-up co-selection.
\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
instead of by the carbon source provided to the communities.
Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
yield a poor correspondence between dominant pairwise competition and coalescence outcomes
(left panel, $R^2=0.07$, $p>0.05$)
and vice-versa
(right panel, $R^2=0.37$, $p<10^{-4}$).}
\label{fig3}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...}
\label{fig4}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig5.pdf}
\caption{\textbf{Community hierarchies determine coalescence outcomes.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...
\textbf{b.} Blablabla... ($p=0.083$).}
\label{fig5}
\end{figure}
